---
title: "DEG_final"
author: "Jacopo Umberto Verga"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# DGE with Seurat

## Multiple Myeloma

THe test with different alogirthms identified seurat as best method. Now I am going to run the DGE analysis with min.pct of cells as 0.25, logistic regression, correcting for the different batches:

```{r message=FALSE, warning=FALSE}
library(Seurat)
library(tidyverse)
library(clusterProfiler)
library(org.Hs.eg.db) 
library(enrichplot)
library(GOSemSim)
library(VennDiagram)

dat <- readRDS("mm_nk.Rds")
dat = SetIdent(dat, value = "new_label")
dat
```


### normalise the data

```{r message=FALSE, warning=FALSE}
dat <- NormalizeData(dat,  normalization.method = "LogNormalize")
```
### DGE analysis

```{r message=FALSE, warning=FALSE}
dat_logreg <- FindAllMarkers(dat, test.use = "LR", min.pct = 0.25,logfc.threshold=0, latent.vars = "batch")
#dat_logreg2 <- filter_DE(dat_logreg, 0)

#extract genes
mm_logreg_exh <- dat_logreg[dat_logreg$cluster == "NK exhausted",] %>%
  mutate(fdr = p.adjust(p_val, method = "fdr")) %>%
  filter(fdr < 0.05)
  
mm_logreg_res <- dat_logreg[dat_logreg$cluster == "NK resident",] %>%
  mutate(fdr = p.adjust(p_val, method = "fdr")) %>%
  filter(fdr < 0.05)

print("DE genes NK exhausted")
dim(mm_logreg_exh)
print("DE genes NK resident")
dim(mm_logreg_res)
```


### GO enrichment analysis

```{r}


GO <- enrichGO(mm_logreg_exh$gene, org.Hs.eg.db, 
               keyType = "SYMBOL", ont="BP",
               pAdjustMethod = "fdr",
               readable="False")

GO <- simplify(GO, cutoff=0.7, by="p.adjust", select_fun = min,measure = "Wang" )

mmGOresult <- GO@result %>%
  filter(p.adjust < 0.05)
dim(mmGOresult)

dotplot(GO)
goplot(GO)


```

### Save files

```{r}
write_csv(mm_logreg_exh, "mm_exhausted.csv")
write_csv(mm_logreg_res, "mm_residemt.csv")
write_csv(mmGOresult, "mm_GO.csv")
```


```{r}
dat <- readRDS("healthy_nk.Rds")
dat = SetIdent(dat, value = "new_label")
dat

```

# normalise the data

```{r message=FALSE, warning=FALSE}
dat <- NormalizeData(dat,  normalization.method = "LogNormalize")
```


```{r message=FALSE, warning=FALSE}
dat_logreg <- FindAllMarkers(dat, test.use = "LR", min.pct = 0.25,logfc.threshold=0, latent.vars = "batch")
#dat_logreg2 <- filter_DE(dat_logreg, 0)

#extract genes
ht_logreg_exh <- dat_logreg[dat_logreg$cluster == "NK exhausted",] %>%
  mutate(fdr = p.adjust(p_val, method = "fdr")) %>%
  filter(fdr < 0.05)

ht_logreg_res <- dat_logreg[dat_logreg$cluster == "NK resident",] %>%
  mutate(fdr = p.adjust(p_val, method = "fdr")) %>%
  filter(fdr < 0.05)

print("DE genes NK exhausted")
dim(ht_logreg_exh)
print("DE genes NK resident")
dim(ht_logreg_res)

```

### GO enrichment analysis

```{r}

genes <- ht_logreg_exh$gene

GO <- enrichGO(ht_logreg_exh$gene, org.Hs.eg.db, 
               keyType = "SYMBOL", ont="BP",
               pAdjustMethod = "fdr",
               readable="False")

GO <- simplify(GO, cutoff=0.7, by="p.adjust", select_fun = min,measure = "Wang" )

htGOresult <- GO@result %>%
  filter(p.adjust < 0.05)
dim(htGOresult)

dotplot(GO)
goplot(GO)


```

```{r}
write_csv(ht_logreg_exh, "healthy_exhausted.csv")
write_csv(ht_logreg_res, "healthy_residemt.csv")
write_csv(htGOresult, "healthy_GO.csv")

```


# Semantic similarity analysis

## Genes

shared or not:

```{r}
mm <- mm_logreg_exh$gene
ht <- ht_logreg_exh$gene

grid.newpage()
venn_object <- venn.diagram(
        x = list(mm, ht),
        category.names = c("MM", "Healthy"),
        main = "DEG between exhausted cells MM and healthy samples",
        main.cex =2,
        cex=2,
        cat.cex=1.5,
        label.col=3,
        filename = NULL,
        disable.logging = T
        )
grid.draw(venn_object)
```



```{r}
hsGO <- godata(org.Hs.eg.db, keytype= "SYMBOL", ont="BP")

# Genes unique for mm

mm_unique <- setdiff(mm,ht)
ht_unique <- setdiff(ht,mm)

clusterSim(mm_unique, ht_unique, semData=hsGO, measure="Wang", combine="BMA")
```

## GOs

```{r}
mm <- mmGOresult$ID
ht <- htGOresult$ID

grid.newpage()
venn_object <- venn.diagram(
        x = list(mm, ht),
        category.names = c("GO:BP mm-exh", "GO:BP healthy-exh"),
        main.cex =2,
        cex=2,
        cat.cex=1.5,
        label.col=3,
        filename = NULL,
        disable.logging = T
        )
grid.draw(venn_object)

# Genes unique for mm

mm_unique <- setdiff(mm,ht)
ht_unique <- setdiff(ht,mm)

mgoSim(mm_unique, ht_unique, semData=hsGO, measure="Wang", combine="BMA")
```
